A tool set for random number generation on GPUs in R

### 2021-11-16 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2973 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.155   0.000   0.585
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   7.388   0.432   7.814
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

Below is an example that we produce a 2 by 2 matrix of exponential random number with expectation 1.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
r_matrix <- clrng::rexp(c(2,2), rate=1, Nglobal=c(1,4), type="double")
as.matrix(r_matrix)
##           [,1]      [,2]
## [1,] 0.9340107 5.0051762
## [2,] 3.4506078 0.8855801
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

\caption{\label{tab:monthdata}Month\label{tab:month}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
  & Ane & Men & Cya & Her & Omp & Gas & Lim & Cle & Pal & Down & Chro & Hypo\\
\midrule
Jan & 29 & 55 & 172 & 46 & 39 & 73 & 48 & 183 & 77 & 103 & 102 & 174\\
Feb & 25 & 45 & 175 & 35 & 31 & 55 & 34 & 142 & 81 & 115 & 100 & 180\\
Mar & 31 & 48 & 182 & 41 & 47 & 72 & 40 & 200 & 86 & 90 & 96 & 180\\
Apr & 34 & 45 & 186 & 36 & 32 & 75 & 42 & 173 & 56 & 87 & 90 & 193\\
May & 33 & 40 & 187 & 46 & 24 & 80 & 35 & 180 & 75 & 91 & 100 & 197\\
Jun & 34 & 48 & 189 & 35 & 33 & 75 & 45 & 154 & 74 & 102 & 100 & 182\\
\bottomrule
\end{tabular}
\end{table}

@@

month_gpu<-vclMatrix(month,type="integer")
system.time(result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64)))
##    user  system elapsed 
##   1.057   0.004   2.596
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}
We got 409524 cases whose test statistics are below the observed threshold based on an actual number of 1015808 simulations executed, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
system.time(result_monthcpu<-stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  18.606   0.027  18.626
result_monthcpu$p.value
## [1] 0.4036733
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\caption{\label{tab:weekdata}Week\label{tab:week}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
  & Ane & Men & Cya & Her & Omp & Gas & Lim & Cle & Pal & Down & Chro & Hypo\\
\midrule
Mon & 30 & 34 & 173 & 37 & 23 & 80 & 49 & 191 & 83 & 122 & 109 & 216\\
Tue & 60 & 121 & 383 & 80 & 83 & 131 & 71 & 349 & 146 & 164 & 168 & 352\\
Wed & 51 & 106 & 417 & 92 & 73 & 145 & 72 & 333 & 136 & 179 & 196 & 351\\
Thu & 60 & 86 & 362 & 69 & 74 & 120 & 85 & 326 & 132 & 220 & 187 & 359\\
Fri & 52 & 94 & 347 & 87 & 59 & 123 & 68 & 323 & 145 & 170 & 166 & 345\\
Sat & 52 & 63 & 323 & 67 & 64 & 135 & 73 & 316 & 170 & 189 & 188 & 357\\
Sun & 49 & 51 & 211 & 40 & 32 & 96 & 69 & 216 & 108 & 143 & 130 & 258\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -clrng::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_weekgpu<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   0.925   0.012  10.653
result_weekgpu$simNum
## [1] 10010624
result_weekgpu$counts
## [1] 1193
result_weekgpu$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
system.time(result_weekcpu<-fisher.test(week,simulate.p.value = TRUE,B=10010624))
##    user  system elapsed 
## 117.213   0.152 117.289
result_weekcpu$p.value
## [1] 0.0001249672
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.156   0.000   0.586
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.951   0.444   7.390
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

Below is an example that we produce a 2 by 2 matrix of exponential random number with expectation 1.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
r_matrix <- clrng::rexp(c(2,2), rate=1, Nglobal=c(1,4), type="double")
as.matrix(r_matrix)
##           [,1]        [,2]
## [1,] 0.2627643 0.005952261
## [2,] 0.4535731 0.186351621
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{clrng::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
  \item \code{verbose}: print extra information if $\text{verbose} > 1$. Default is FALSE.
\end{itemize}

@@

\caption{\label{tab:monthdata}Month\label{tab:month}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
  & Ane & Men & Cya & Her & Omp & Gas & Lim & Cle & Pal & Dow & Chr & Hyp\\
\midrule
Jan & 29 & 55 & 172 & 46 & 39 & 73 & 48 & 183 & 77 & 103 & 102 & 174\\
Feb & 25 & 45 & 175 & 35 & 31 & 55 & 34 & 142 & 81 & 115 & 100 & 180\\
Mar & 31 & 48 & 182 & 41 & 47 & 72 & 40 & 200 & 86 & 90 & 96 & 180\\
Apr & 34 & 45 & 186 & 36 & 32 & 75 & 42 & 173 & 56 & 87 & 90 & 193\\
May & 33 & 40 & 187 & 46 & 24 & 80 & 35 & 180 & 75 & 91 & 100 & 197\\
Jun & 34 & 48 & 189 & 35 & 33 & 75 & 45 & 154 & 74 & 102 & 100 & 182\\
\bottomrule
\end{tabular}
\end{table}

@@

month_gpu<-vclMatrix(month,type="integer")
system.time(result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64)))
##    user  system elapsed 
##   1.023   0.028   2.587
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}
We got 409524 cases whose test statistics are below the observed threshold based on an actual number of 1015808 simulations, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
system.time(result_monthcpu<-stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  18.341   0.008  18.336
result_monthcpu$p.value
## [1] 0.4039637
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\caption{\label{tab:weekdata}Week\label{tab:week}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
  & Ane & Men & Cya & Her & Omp & Gas & Lim & Cle & Pal & Dow & Chr & Hyp\\
\midrule
Mon & 30 & 34 & 173 & 37 & 23 & 80 & 49 & 191 & 83 & 122 & 109 & 216\\
Tue & 60 & 121 & 383 & 80 & 83 & 131 & 71 & 349 & 146 & 164 & 168 & 352\\
Wed & 51 & 106 & 417 & 92 & 73 & 145 & 72 & 333 & 136 & 179 & 196 & 351\\
Thu & 60 & 86 & 362 & 69 & 74 & 120 & 85 & 326 & 132 & 220 & 187 & 359\\
Fri & 52 & 94 & 347 & 87 & 59 & 123 & 68 & 323 & 145 & 170 & 166 & 345\\
Sat & 52 & 63 & 323 & 67 & 64 & 135 & 73 & 316 & 170 & 189 & 188 & 357\\
Sun & 49 & 51 & 211 & 40 & 32 & 96 & 69 & 216 & 108 & 143 & 130 & 258\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -clrng::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_weekgpu<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   0.869   0.013  10.601
result_weekgpu$simNum
## [1] 10010624
result_weekgpu$counts
## [1] 1193
result_weekgpu$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
system.time(result_weekcpu<-fisher.test(week,simulate.p.value = TRUE,B=10010624))
##    user  system elapsed 
## 115.767   0.111 115.784
result_weekcpu$p.value
## [1] 0.0001297621
\end{verbatim}
\end{kframe}
\end{knitrout}
